Next Trippy Structures\thanks{Supported by organization x.}

### 2019-02-12 paper.tex
Avoid referring to CTR structures by indices

\subsubsection{Network constrained trajectory indexing}
There are also numerous approaches that use R-Tree based indexes for trajectories that are constrained to an underlying network, aiming to decrease the redundancy in the representation by separating the representation into levels. Examples of such indexes include the {\em FNR-Tree}~\cite{DBLP:conf/ssd/Frentzos03}, where the network elements are indexed with a 2D R-Tree and 1D R-Trees to index time intervals of moving objects along the edges of the network. Some of the limitations of the FNR-Tree are addressed in \cite{DBLP:journals/geoinformatica/AlmeidaG05}, where the authors propose the {\em MON-Tree}, where the moving objects are indexes in two dimensional R-Trees (the dimensions being edge position vs time). More recent alternatives also integrate {\em B$^+$-Trees}, which has proven to be more space and time efficient for indexing the temporal dimension as proven by the {\em TMN-Tree}~\cite{chang2010tmn}. Refer to \cite{john2017performance} for a quick comparison of these R-Tree based indexes.

@@

\marginpar{We should probably cite our previous work too, right?}
A recent compact data structure names {\em CiNCT} has been proposed in \cite{koide2018cinct}, where trajectories are encoded into a string, that is used to build an FM-index \cite{DBLP:conf/focs/FerraginaM00} with a Huffman-shaped Wavelet Tree \cite{ferragina2009compressed}. To further save space, the string is constructed with relative movement labels instead of absolute edges, with an auxiliary structure that represents a network graph built from the input trajectories themselves. While this structure excels at pattern-matching and path extraction in vehicular networks (such as the streets of a city), it cannot be applied to our problems in public transportation networks.

\marginpar{\cite{rivera2018faster} is just 2 bitmaps to answer queries on time intervals, not directly related to trajectories despite what the title suggests. Should we cite it just to be nice? If so, then how?}
% \cite{rivera2018faster}, which introduces two bitvectors to efficiently answer time interval queries and us...

@@

That is, $B_1$ contains all the bits $a_1 \forall a \in S$, in the same order as in S. After that, $B_2$ contains the $a_2$ bits from those $a$ such as $a_1 = 0$ first, followed by those with $a_1 = 1$. That way, all the position of each $a_i$ in $B_i$ depends first by the value of $a_{i-1}$ and later by their position in $B_{i-1}$. Refer to the structures (ii) and (iii) in Figure~\ref{fig:example_ctr} for examples of one and two-leveled WM. Additionally, there exists a conceptual \textit{leaf level}, that is formed by the final position of every $a$ after applying the recursion on the last level. Because the sorting of each level depends on the values of the previous most significant bits, in the leaf level all symbols are sorted by their reverse representation in bits.

@@

\begin{figure}[hbt!]
\includegraphics[width=\textwidth]{example_trips.eps}
\caption{An example of five trips represented on \ctr with the optimizations for (ii) and (iii), and sections for each stop delimited by dotted lines}
\label{fig:example_ctr}
\end{figure}

We use three complementary structures to represent each component of the sequence, as shown in Figure~\ref{fig:example_ctr}:
\begin{enumerate}[(i)]
    \item An adapted Compressed Suffix Array (CSA) over the stop identifiers of all trips, concatenated into a string with additional terminator symbols $\$$ appended at the end of each trip. While in the final CSA we assign these $\$$ a lexicographical value of 0 (thus smaller than any stop identifier), we assign them different values during the construction of the suffix array (A) to ensure that the entries for $\$$ in A maintain the same order as in the original text. Finally, we make a modification on $\Psi$ to make the entries of each $\$$ point to the start of its own trip instead of the next one. These two modifications are proven necessary for our implemented queries, at the expense of losing some of the properties of a classic CSA.
    \item WML: Aligned to the entries of (i) there is a Wavelet Matrix (WM) for the line identifiers of each stop. Aligned to the $\$$ section we duplicate the starting lines of each trip. As a trivial optimization, we build a separate WM for every stop, allowing us to save space due to the fact that a single stop does not usually belong to many lines, thus the average height of these WM is no larger (and usually smaller) than the height of a single WM.
    \item WVJ: A WM of jcodes aligned to the leaves of (ii). Note that this makes this structure dependant on (ii), which is coherent with the fact that journey codes themselves are relative to the line identifier. In case (ii) implements the optimization described, the entries of the WM must also be rearranged to match the delimited stops.
\end{enumerate}

As \ctr is a complete representation of the collection of trips, we are able to extract any trip using the structures described. \marginpar{The idea here is that explaining how to extract will help understand the more complex FromXtoY query later} The algorithm to extract a trip is shown in the Algorithm~\ref{alg:extract}, where \FuncSty{Rank} and \FuncSty{Select} operate over the bitvector $D$ from (i), \FuncSty{WML}{(s$_a$)} is the WM corresponding to the stop \DataSty{s$_a$} in the optimized version of (ii) and \FuncSty{TrackDown} returns the leaf index of a WM given a root index. In a practical implementation, it is not needed to access (ii) and (iii) for the line identifier and jcode of the last stop of the trip, as they will always match the previous ones.

@@

 \caption{Extracting the trip \DataSty{i} from \ctr where \FuncSty{CSA}, \FuncSty{WML} and \FuncSty{WMJ} are structures (i), (ii) and (iii), respectively}
 \label{alg:extract}
\end{algorithm}

@@

 \caption{Querying for all features on \ctr where \FuncSty{CSA}, \FuncSty{WML} and \FuncSty{WMJ} are structures (i), (ii) and (iii), respectively}
 \label{alg:xy}
\end{algorithm}

We will now proceed to explain the Algorithm~\ref{alg:xy} line by line:
\begin{itemize}
    \item In \textbf{lines 2-3}, we query the \FuncSty{CSA} in (i) for the pattern consisting of the destination stop \DataSty{s$_z$}, followed by a 0 (the code for the $\$$ symbol) and finally the origin stop \DataSty{s$_a$}. This results in a range of entries within the section of \DataSty{s$_z$} that belong to our queried trips, because the trips were made circular in $\Psi$, so each $\$$ points to the beginning of its own trip. If we were not interested in restricting lines nor time, the function would end here, returning \DataSty{right}-\DataSty{left}.

@@

    \item In \textbf{line 7} we query (ii) in the $\$$ section, within the range previously obtained, for the queried starting line \DataSty{l$_a$}, obtaining the range of its occurrences in the last level, delimited by \DataSty{a} and \DataSty{z}. Note that if the \ctr was constructed without the optimization that separates (ii) in sections, this line would be exactly the same, save for the query being on a \DataSty{WML} that would encode the whole sequence of lines instead of just the \DataSty{WML}{(0)} for $\$$.

@@

    \item In \textbf{line 9} we operate over a range of (iii) that encodes a non-decreasing sequence of jcodes, given that within the same origin stop, final stop and starting line, the $\$$ were sorted by the starting journey code of their trips. This allows us to use a specially modified version of \FuncSty{GetRange} that is able to return the range of indexes on the first level of the WM instead of the last one. This is equivalent to finding the lower and upper bounds for the range \DataSty{j$_a$}$..$\DataSty{j$_z$} via binary searches in \FuncSty{WMJ} from \DataSty{a} to \DataSty{z}, but much more efficient.

    \item In \textbf{lines 10-12} we use the \FuncSty{TrackUp} operation, which is the inverse of \FuncSty{TrackDown}: it returns the root index given a leaf index of a WM. In this case, as the \FuncSty{a} and \FuncSty{z} we obtained in the previous step are also indexes in the last level of the lines WM, we use \FuncSty{TrackUp} to translate that range of indexes to the root of (ii) and therefore to the entries of (i) as well. Note that this is a range inside the $\$$ section for trips with the same origin stop, final stop and starting line that includes all the trips for jcodes that span from \DataSty{j$_a$} to \DataSty{j$_z$}. The properties of (i) also ensure that this range maintains the same size after translation and it can also be directly translated to a range in the \DataSty{s$_z$}.

    \item In \textbf{line 13} we obtain the starting index of the section for the stop \DataSty{s$_z$} with a \FuncSty{Select} operation over the bitvector D from (i). This is necessary for operating on (ii) and (iii) to restrict the line and journeys for the final stop.

    \item In \textbf{line 14} the range \DataSty{a'}$..$\DataSty{z'} is trivially translated to the section for \DataSty{s$_z$} where we query (ii) to obtain the subrange for the line \DataSty{l$_z$}. Remember that \DataSty{WML}{(s$_z$)} represents only the lines for \DataSty{s$_z$}, therefore both the translated indexes and the resulting subrange indexes are relative and must be adjusted by \DataSty{offset}. This would have not been necessary if the optimization was not implemented, and absolute indexes would have been used.

@@

    \item In \textbf{line 16} we return the number of entries of (iii) between \DataSty{j$_a$} and \DataSty{j$_z$} within our final subrange. The function \FuncSty{GetCount} does not report any range boundaries as \FuncSty{GetRange} does, but simply returns the number of occurrences. Of course, the time complexity of this operation is still proportional to the height of the WM even though the matching entries can be scattered (not delimiting a contiguous range).
\end{itemize}

It is trivial to reuse the Algorithm~\ref{alg:xy} to answer queries with less constrains. For example, if we were only interested in restricting the starting line, we could return \DataSty{z}-\DataSty{a} after line 7. Or if we only wanted to restrict the ending line and time, we can do it by skipping the lines 4-12, and using directly \FuncSty{GetRange}{(\FuncSty{WML}{(s$_z$)},left-offset,right-offset,l$_z$)} in line 14. The complexity would increase if we restricted by a time window but not by lines, as would need to iterate through all possible lines for \DataSty{s$_a$} and \DataSty{s$_z$} to obtain the jcodes for each line and perform these operations on (ii) and (iii). Fortunately, the number of lines that a single stop can belong to tends to be rather small in practice, thus with a careful implementation that avoids repeating computation the performance of such query scales well, as will be shown later in Section~\ref{sec:exp}. Finally, to obtain only the trips that started on a given stop we would simply need to set \DataSty{pattern} to $\{0,s_a\}$ in line 2, or alternatively to $\{s_z,0\}$ for the final stop, and skipping the operations on the sections of \DataSty{s$_z$} or $\$$, respectively.

>>>>

\subsubsection{Network constrained trajectory indexing}
There are also numerous approaches that use R-Tree based indexes for trajectories that are constrained to an underlying network, aiming to decrease the redundancy in the representation by separating the representation into levels. Examples of such indexes include the {\em FNR-Tree}~\cite{DBLP:conf/ssd/Frentzos03}, where the network elements are indexed with a 2D R-Tree and 1D R-Trees to index time intervals of moving objects along the edges of the network. Some of the limitations of the FNR-Tree are addressed in \cite{DBLP:journals/geoinformatica/AlmeidaG05}, where the authors propose the {\em MON-Tree}, where the moving objects are indexes in two dimensional R-Trees (the dimensions being edge position vs time). More recent alternatives also integrate {\em B$^+$-Trees}, which has proven to be more space and time efficient for indexing the temporal dimension as proven by the {\em TMN-Tree}~\cite{chang2010tmn}. Alternatively, in \cite{rivera2018faster} a compact representation of time intervals is proposed using two bitvectors, that can be combined with these R-Tree based indexes to increase the efficiency for temporal filtering. Refer to \cite{john2017performance} for a quick comparison of these R-Tree based indexes.

@@

A recent compact data structure names {\em CiNCT} has been proposed in \cite{koide2018cinct}, where trajectories are encoded into a string, that is used to build an FM-index \cite{DBLP:conf/focs/FerraginaM00} with a Huffman-shaped Wavelet Tree \cite{ferragina2009compressed}. To further save space, the string is constructed with relative movement labels instead of absolute edges, with an auxiliary structure that represents a network graph built from the input trajectories themselves. While this structure excels at pattern-matching and path extraction in vehicular networks (such as the streets of a city), it cannot be applied to our problems in public transportation networks.

@@

That is, $B_1$ contains all the bits $a_1 \forall a \in S$, in the same order as in S. After that, $B_2$ contains the $a_2$ bits from those $a$ such as $a_1 = 0$ first, followed by those with $a_1 = 1$. That way, all the position of each $a_i$ in $B_i$ depends first by the value of $a_{i-1}$ and later by their position in $B_{i-1}$. Refer to the structures \texttt{WML} and \texttt{WMJ} in Figure~\ref{fig:example_ctr} for examples of one and two-leveled WM. Additionally, there exists a conceptual \textit{leaf level}, that is formed by the final position of every $a$ after applying the recursion on the last level. Because the sorting of each level depends on the values of the previous most significant bits, in the leaf level all symbols are sorted by their reverse representation in bits.

@@

\begin{figure}[hbt!]
\includegraphics[width=\textwidth]{example_trips.eps}
\caption{An example of five trips represented on \ctr with the optimizations for \texttt{WML} and \texttt{WMJ}, and sections for each stop delimited by dotted lines}
\label{fig:example_ctr}
\end{figure}

We use three complementary structures to represent each component of the sequence, as shown in Figure~\ref{fig:example_ctr}:
\begin{enumerate}[(i)]
    \item An adapted Compressed Suffix Array (\texttt{CSA}) over the stop identifiers of all trips, concatenated into a string with additional terminator symbols $\$$ appended at the end of each trip. While in the final CSA we assign these $\$$ a lexicographical value of 0 (thus smaller than any stop identifier), we assign them different values during the construction of the suffix array (A) to ensure that the entries for $\$$ in A maintain the same order as in the original text. Finally, we make a modification on $\Psi$ to make the entries of each $\$$ point to the start of its own trip instead of the next one. These two modifications are proven necessary for our implemented queries, at the expense of losing some of the properties of a classic CSA.
    \item \texttt{WML}: Aligned to the entries of (i) there is a Wavelet Matrix (WM) for the line identifiers of each stop. Aligned to the $\$$ section we duplicate the starting lines of each trip. As a trivial optimization, we build a separate WM for every stop, allowing us to save space due to the fact that a single stop does not usually belong to many lines, thus the average height of these WM is no larger (and usually smaller) than the height of a single WM.
    \item \texttt{WMJ}: A WM of jcodes aligned to the leaves of (ii). Note that this makes this structure dependant on (ii), which is coherent with the fact that journey codes themselves are relative to the line identifier. In case (ii) implements the optimization described, the entries of the WM must also be rearranged to match the delimited stops.
\end{enumerate}

As \ctr is a complete representation of the collection of trips, we are able to extract any trip using the structures described. \marginpar{The idea here is that explaining how to extract will help understand the more complex FromXtoY query later} The algorithm to extract a trip is shown in the Algorithm~\ref{alg:extract}, where \FuncSty{Rank} and \FuncSty{Select} operate over the bitvector $D$ from our \texttt{CSA}, \FuncSty{WML}{(s$_a$)} is the WM corresponding to the stop \DataSty{s$_a$} in the optimized version of \texttt{WML} and \FuncSty{TrackDown} returns the leaf index of a WM given a root index. In a practical implementation, it is not needed to access \texttt{WML} and \texttt{WMJ} for the line identifier and jcode of the last stop of the trip, as they will always match the previous ones.

@@

 \caption{Extracting the trip \DataSty{i} from \ctr where \FuncSty{CSA}, \FuncSty{WML} and \FuncSty{WMJ} are the structures previously described in (i), (ii) and (iii), respectively}
 \label{alg:extract}
\end{algorithm}

@@

 \caption{Querying for all features on \ctr}
 \label{alg:xy}
\end{algorithm}

We will now proceed to explain the Algorithm~\ref{alg:xy} line by line:
\begin{itemize}
    \item In \textbf{lines 2-3}, we query our \FuncSty{CSA} for the pattern consisting of the destination stop \DataSty{s$_z$}, followed by a 0 (the code for the $\$$ symbol) and finally the origin stop \DataSty{s$_a$}. This results in a range of entries within the section of \DataSty{s$_z$} that belong to our queried trips, because the trips were made circular in $\Psi$, so each $\$$ points to the beginning of its own trip. If we were not interested in restricting lines nor time, the function would end here, returning \DataSty{right}-\DataSty{left}.

@@

    \item In \textbf{line 7} we query \texttt{WML} in the $\$$ section, within the range previously obtained, for the queried starting line \DataSty{l$_a$}, obtaining the range of its occurrences in the last level, delimited by \DataSty{a} and \DataSty{z}. Note that if the \ctr was constructed without the optimization that separates \texttt{WML} in sections, this line would be exactly the same, save for the query being on a \DataSty{WML} that would encode the whole sequence of lines instead of just the \DataSty{WML}{(0)} for $\$$.

@@

    \item In \textbf{line 9} we operate over a range of \texttt{WMJ} that encodes a non-decreasing sequence of jcodes, given that within the same origin stop, final stop and starting line, the $\$$ were sorted by the starting journey code of their trips. This allows us to use a specially modified version of \FuncSty{GetRange} that is able to return the range of indexes on the first level of the WM instead of the last one. This is equivalent to finding the lower and upper bounds for the range \DataSty{j$_a$}$..$\DataSty{j$_z$} via binary searches in \FuncSty{WMJ} from \DataSty{a} to \DataSty{z}, but much more efficient.

    \item In \textbf{lines 10-12} we use the \FuncSty{TrackUp} operation, which is the inverse of \FuncSty{TrackDown}: it returns the root index given a leaf index of a WM. In this case, as the \FuncSty{a} and \FuncSty{z} we obtained in the previous step are also indexes in the last level of the lines WM, we use \FuncSty{TrackUp} to translate that range of indexes to the root of \texttt{WML} and therefore to the entries of the \texttt{CSA} as well. Note that this is a range inside the $\$$ section for trips with the same origin stop, final stop and starting line that includes all the trips for jcodes that span from \DataSty{j$_a$} to \DataSty{j$_z$}. The properties of our adapted \texttt{CSA} also ensure that this range maintains the same size after translation and it can also be directly translated to a range in the \DataSty{s$_z$}.

    \item In \textbf{line 13} we obtain the starting index of the section for the stop \DataSty{s$_z$} with a \FuncSty{Select} operation over the bitvector D from \texttt{CSA}. This is necessary for operating on \texttt{WML} and \texttt{WMJ} to restrict the line and journeys for the final stop.

    \item In \textbf{line 14} the range \DataSty{a'}$..$\DataSty{z'} is trivially translated to the section for \DataSty{s$_z$} where we query \texttt{WML} to obtain the subrange for the line \DataSty{l$_z$}. Remember that \DataSty{WML}{(s$_z$)} represents only the lines for \DataSty{s$_z$}, therefore both the translated indexes and the resulting subrange indexes are relative and must be adjusted by \DataSty{offset}. This would have not been necessary if the optimization was not implemented, and absolute indexes would have been used.

@@

    \item In \textbf{line 16} we return the number of entries of \texttt{WMJ} between \DataSty{j$_a$} and \DataSty{j$_z$} within our final subrange. The function \FuncSty{GetCount} does not report any range boundaries as \FuncSty{GetRange} does, but simply returns the number of occurrences. Of course, the time complexity of this operation is still proportional to the height of the WM even though the matching entries can be scattered (not delimiting a contiguous range).
\end{itemize}

It is trivial to reuse the Algorithm~\ref{alg:xy} to answer queries with less constrains. For example, if we were only interested in restricting the starting line, we could return \DataSty{z}-\DataSty{a} after line 7. Or if we only wanted to restrict the ending line and time, we can do it by skipping the lines 4-12, and using directly \FuncSty{GetRange}{(\FuncSty{WML}{(s$_z$)},left-offset,right-offset,l$_z$)} in line 14. The complexity would increase if we restricted by a time window but not by lines, as would need to iterate through all possible lines for \DataSty{s$_a$} and \DataSty{s$_z$} to obtain the jcodes for each line and perform these operations on \texttt{WML} and \texttt{WMJ}. Fortunately, the number of lines that a single stop can belong to tends to be rather small in practice, thus with a careful implementation that avoids repeating computation the performance of such query scales well, as will be shown later in Section~\ref{sec:exp}. Finally, to obtain only the trips that started on a given stop we would simply need to set \DataSty{pattern} to $\{0,s_a\}$ in line 2, or alternatively to $\{s_z,0\}$ for the final stop, and skipping the operations on the sections of \DataSty{s$_z$} or $\$$, respectively.
